KidsAIR initial data work

### 2019-11-26 data_cleaning_initial.Rmd
further data processing/cleaning

# Rename and create new variables
kids_pm <- pm_initial %>% 
  # separate time from incorrect/associated date
  separate(time, into = c("trash", "time"), sep = " ") %>% 
  # format/rename variables
  mutate(id_home = as.factor(homeid),
         pm_date = ymd(date),
         pm_time = hms(time),
         pm = as.numeric(aerosol)) %>% 
  # new datetime variable
  unite(pm_datetime, c("date", "time"), sep = " ", remove = FALSE) %>% 
  select(id_home, area, pm_date, pm_time, pm_datetime, pm) %>% 
  arrange(id_home, pm_date) %>% 
  group_by(id_home) %>% 
  # new variables summarizing pm2.5 within each id_home
  mutate(pm_total_observations = n(),
         pm_total_summed = sum(pm),
         min_datetime = first(pm_datetime),
         max_datetime = last(pm_datetime),
         min_datetime = ymd_hms(min_datetime),
         max_datetime = ymd_hms(max_datetime),
         diff_datetime = interval(min_datetime, max_datetime),
         pm_sampling_days = diff_datetime/86400,
         pm_24hr_mean = pm_total_summed/pm_sampling_days,
         pm_rolling_24hr = rollsum(pm, 1440, fill = "extend", align = "right")) %>% 
  select(-min_datetime, -max_datetime, -diff_datetime) %>% 
  ungroup()

@@

# Load, format, and save fuel moisture content data
```{r}
moisture_initial_wmt <- read_xlsx("Input/wmt_kids_moisture_20191122.xlsx")

kids_moisture <- moisture_initial_wmt %>% 
  mutate(id = as.factor(ID), # ask about what these IDs are 
         id_home = as.factor(HomeWinterID),
         moisture_date = ymd(CompletedDate),
         moisture_1 = as.numeric(Moisture1),
         moisture_2 = as.numeric(Moisture2),
         moisture_3 = as.numeric(Moisture3),
         moisture_ave = as.numeric(MoistureAve),
         moisture_missing = as.character(WhatMissing),
         moisture_incomplete = as.factor(Incomplete),
         moisture_ignore_missing = as.factor(IgnoreMissingData),
         moisture_ignore_reason = as.character(IgnoreReason),
         moisture_tech = as.character(Technician),
         moisture_split = as.factor(Split),
         area = "WMT") %>% 
  select(id, id_home, area, moisture_date, moisture_1, moisture_2, moisture_3, 
         moisture_ave, moisture_missing, moisture_incomplete, moisture_ignore_missing,
         moisture_ignore_reason, moisture_tech, moisture_split)

# Save pm dataset as RDS and CSV
write_rds(kids_moisture, "Output/kids_moisture.rds")
write_csv(kids_moisture, "Output/kids_moisture.csv")
```

>>>>

# Rename and create new variables
kids_pm <- pm_initial %>% 
  # separate time from incorrect/associated date
  separate(time, into = c("trash", "time"), sep = " ") %>% 
  # format/rename variables
  mutate(home_winter_id = as.factor(homeid),
         pm_date = ymd(date),
         pm_time = hms(time),
         pm = as.numeric(aerosol)) %>% 
  # new datetime variable
  unite(pm_datetime, c("date", "time"), sep = " ", remove = FALSE) %>% 
  select(home_winter_id, area, pm_date, pm_time, pm_datetime, pm) %>% 
  arrange(home_winter_id, pm_date) %>% 
  group_by(home_winter_id) %>% 
  # new variables summarizing pm2.5 within each home_winter_id
  mutate(pm_total_observations = n(),
         pm_total_summed = sum(pm),
         min_datetime = first(pm_datetime),
         max_datetime = last(pm_datetime),
         min_datetime = ymd_hms(min_datetime),
         max_datetime = ymd_hms(max_datetime),
         diff_datetime = interval(min_datetime, max_datetime),
         pm_total_days = diff_datetime/86400,
         pm_24hr_mean = pm_total_summed/pm_total_days,
         pm_rolling_24hr = rollsum(pm, 1440, fill = "extend", align = "right"),
         obs = row_number(),
         sampling_day = if_else(obs < 1441, 1, 0),
         sampling_day = if_else(obs < 2881 & obs > 1440, 2, sampling_day),
         sampling_day = if_else(obs < 4321 & obs > 2880, 3, sampling_day),
         sampling_day = if_else(obs < 5761 & obs > 4320, 4, sampling_day),
         sampling_day = if_else(obs < 7201 & obs > 5760, 5, sampling_day),
         sampling_day = if_else(obs < 8641 & obs > 7200, 6, sampling_day),
         sampling_day = if_else(obs < 10081 & obs > 8640, 7, sampling_day),
         sampling_day = if_else(obs > 10080, 8, sampling_day)) %>%
  group_by(home_winter_id, sampling_day) %>% 
  mutate(pm_24hr_daily = sum(pm),
         pm_24hr_daily_weighted = pm_24hr_daily*(1440/n()),
         pm_sampling_day = sampling_day) %>% 
  ungroup() %>% 
  select(-min_datetime, -max_datetime, -diff_datetime, -obs, -sampling_day)

@@

# Load, format, and save fuel moisture content data
```{r}
# Load data
moisture_initial_wmt <- read_xlsx("Input/wmt_kids_moisture_20191122.xlsx")

kids_moisture <- moisture_initial_wmt %>% 
  # rename variables
  mutate(moisture_id = as.factor(ID),  
         home_winter_id = as.factor(HomeWinterID),
         moisture_date = ymd(CompletedDate),
         moisture_1 = as.numeric(Moisture1),
         moisture_2 = as.numeric(Moisture2),
         moisture_3 = as.numeric(Moisture3),
         moisture_ave = as.numeric(MoistureAve),
         moisture_missing = as.character(WhatMissing),
         moisture_incomplete = as.factor(Incomplete),
         moisture_ignore_missing = as.factor(IgnoreMissingData),
         moisture_ignore_reason = as.character(IgnoreReason),
         moisture_tech = as.character(Technician),
         moisture_split = as.factor(Split),
         area = "WMT") %>% 
  # replace 0 and -999 values with NA
  replace_with_na(replace = list(moisture_1 = 0, moisture_2 = 0,
                                 moisture_3 = 0, moisture_ave = 0)) %>% 
  replace_with_na(replace = list(moisture_1 = -999, moisture_2 = -999,
                                 moisture_3 = -999)) %>% 
  # gather 3 moisture samples to calculate average
  gather("sample", "value", c("moisture_1":"moisture_3")) %>% 
  group_by(home_winter_id, moisture_date) %>% 
  mutate(moisture_ave = mean(value, na.rm = TRUE)) %>% 
  ungroup() %>% 
  # spread 3 moisture samples back to wide format
  spread(sample, value) %>% 
  select(moisture_id, home_winter_id, area, moisture_date, moisture_1, moisture_2, moisture_3, 
         moisture_ave, moisture_missing, moisture_incomplete, moisture_ignore_missing,
         moisture_ignore_reason, moisture_tech, moisture_split)

# Save as RDS and CSV
write_rds(kids_moisture, "Output/kids_moisture.rds")
write_csv(kids_moisture, "Output/kids_moisture.csv")
```

# Load, format, and save home and child overview data
## These files will link the home ID and child ID with the intervention status
```{r}
# Load files
wmt_child_initial <- read_xlsx("Input/wmt_child_20191126.xlsx")
wmt_home_initial <- read_xlsx("Input/wmt_home_20191126.xlsx")
wmt_homewinter_initial <- read_xlsx("Input/wmt_homewinter_20191126.xlsx")
wmt_parent_demographics_initial <- read_xlsx("Input/wmt_parent_demographics_20191126.xlsx")

# join all files from above
joined1 <- wmt_child_initial %>% 
  mutate(hispanic_child = as.factor(Hispanic),
         race_child = as.factor(Race),
         comments_child = Comments,
         what_missing_child = WhatMissing,
         incomplete_child = Incomplete,
         ignore_missing_child = IgnoreMissingData,
         ignore_reason_child = IgnoreReason) %>% 
  select(-ID, -Hispanic, -Race, -Comments, -WhatMissing, -Incomplete,
         -IgnoreMissingData, -IgnoreReason) %>% 
  left_join(wmt_homewinter_initial, by = "HomeID") %>% 
  left_join(wmt_home_initial, by = "HomeID") %>% 
  left_join(wmt_parent_demographics_initial, by = "HomeWinterID") %>%
  mutate(gender_parent = as.factor(Gender),
         hispanic_parent = as.factor(Hispanic),
         race_parent = as.factor(Race),
         what_missing_parent = WhatMissing,
         incomplete_parent = Incomplete,
         ignore_missing_parent = IgnoreMissingData,
         ignore_reason_parent = IgnoreReason) %>%
  select(-ID, -Hispanic, -Race, -WhatMissing, -Incomplete,
         -IgnoreMissingData, -IgnoreReason, -Gender, -EnterDate) %>%
  rename_all(tolower)

# create dataset of id's and treatment assignment
wmt_ids_linked <- joined1 %>% 
  select(childid, homeid, home, homewinterid, winterid, txcondition) %>% 
  mutate(child_id = as.character(childid),
         home_id = as.character(homeid),
         home_winter_id = as.character(homewinterid),
         winter_id = as.character(winterid),
         treatment = as.factor(txcondition)) %>% 
  select(-childid, -homeid, -homewinterid, -winterid, -txcondition)

# create dataset of demographic data
wmt_demographic_data <- joined1 %>% 
  mutate(child_id = as.character(childid),
         home_id = as.character(homeid),
         age_nov1 = as.factor(agenov1),
         child_assent_date = ymd(childassentdate),
         parent_consent_date = ymd(parentconsentdate),
         home_winter_id = as.character(homewinterid),
         total_residents = as.numeric(totalres),
         total_residents_under5 = as.numeric(underfive),
         ignore_blanks_parent = ignoreblanks,
         income = as.factor(income),
         education = as.factor(education)) %>% 
  select(child_id, home_id, hispanic_child, age_nov1, race_child, child_assent_date,
         parent_consent_date, comments_child, what_missing_child, incomplete_child,
         enrolled, ignore_missing_child, ignore_reason_child,
         home_winter_id, gender_parent, hispanic_parent, race_parent, income,
         education, total_residents, total_residents_under5, incomplete_parent,
         ignore_blanks_parent, what_missing_parent, ignore_missing_parent,
         ignore_reason_parent) %>% 
  distinct(child_id, .keep_all = TRUE)

# create dataset of data on home details
## lower on priority list - see which variables we will use and clean later
wmt_home_details_uncleaned <- wmt_homewinter_initial %>%
  left_join(wmt_home_initial, by = "HomeID") %>% 
  rename_all(tolower)

# Save as RDS and CSV
write_rds(wmt_ids_linked, "Output/wmt_ids_linked.rds")
write_rds(wmt_demographic_data, "Output/wmt_demographic_data.rds")
write_csv(wmt_ids_linked, "Output/wmt_ids_linked.csv")
write_csv(wmt_demographic_data, "Output/wmt_demographic_data.csv")
```
